Density-dependent selection and the limits of relative fitness

### 2018-03-10 manuscript.tex
latest

\subsection*{Solving the variable-density lottery}

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations. The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}. Simulations are 	conducted as follows: $x_1,\ldots,x_G$ values are sampled $U$ times from Poisson distributions with respective means $l_1,\ldots,l_G$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_j c_j x_j$. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}. Dashed lines show the breakdown of the classic lottery model. Parameters: $U=10^5$ and $l_1/l_2=0.1$ is kept fixed while varying the total density $L$.} 
\end{figure}

@@

Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because at low density most territories are claimed without contest, whereas at high density few unoccupied territories are available to be contested. To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
\begin{equation}
\frac{\Delta n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i, \label{eq:bdensitydependence}
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. It is clear $d$-selection is independent of density. On the other hand, the strength of $b$-selection declines with density due to the presence of the $T-N$ factor; in words, the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

@@

By contrast, density trajectories for $c$-selection in the variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density. As a result, Eq.~\eqref{eq:canonical} holds in equilibrium even though $c$-selection is density-dependent. We discuss the non-equilibrium case in [BLAH].

\subsection*{Density-dependent selection on density-regulating traits}

@@

At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$. However, equilibrium-to-equilibrium $b$-sweeps are qualitatively different from $\delta$ sweeps. The reason is that $b$ is not a simple per-capita birth rate. Greater $b$ means more propagules contesting territories, but also more territories being contested. The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a $b$-sweep, even though $N$ increases. Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (deviation from Eq.~\eqref{eq:canonical} in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be quite strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the constant-$s$ model effectively applies exactly for equilibrium $b$-sweeps.

>>>>

\subsection*{Analytical approximation of the variable-density lottery}

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations. The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}. Simulations are 	conducted as follows: $x_i,x_j$ values are sampled $U$ times from Poisson distributions with respective means $l_i,l_j$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_i c_i x_i$. Two types are present, a rare type with $c_j=1.5$, and a common type with $c_i=1$. Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}. Dashed lines show the failure of the classic lottery model at low density. Parameters: $U=10^5$ and $l_j/l_i=0.1$.} 
\end{figure}

@@

Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because at low density most territories are claimed without contest, whereas at high density few unoccupied territories are available to be contested. To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
\begin{equation}
\frac{\Delta n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i, \label{eq:bdensitydependence}
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. It is clear $d$-selection is independent of density. On the other hand, the strength of $b$-selection declines with density because the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

@@

By contrast, density trajectories for $c$-selection in the variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density. As a result, Eq.~\eqref{eq:canonical} holds in equilibrium even though $c$-selection is density-dependent.

\subsection*{Density-regulating traits at equilibrium}

@@

In the variable-density lottery, equilibrium-to-equilibrium $b$-sweeps are qualitatively different from $\delta$ sweeps, because $b$ is not a simple birth rate. Greater $b$ means more propagules contesting territories, but also more territories being contested. The density-dependence of $b$-selection is represented by the term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence}. The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a $b$-sweep, even though $N$ increases. Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (deviation from Eq.~\eqref{eq:canonical} in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be quite strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model effectively applies exactly for equilibrium $b$-sweeps (however, selection is frequency-dependent because of the $1/\overline{b}$ factor).

If selection simultaneously acts on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}). This can produce behavior reminiscent of Fig.~\ref{fig:strengthofselection}.

\begin{figure}
\centering
\includegraphics[scale=0.8]{multiple.pdf}
\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep.}
\end{figure}
